---
title: "Work with dataframe"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## basic characteristics of dataframe
```{r}
library(base)#default loading
rownames()
colnames()
rowSums(x, na.rm=F) 
colSums(x, na.rm=F)
factor(df$column1,ordered=T, levels= c("low", "mid", "high"))
summary(as.factor(df$column1))
```

#piping:select columns, filter rows, add new columns, calculate and summarize data
```{r}
library(dplyr)
arrange(df, A,C,D)
select(df, A:E)
rename(df, A, new_A)
filter(df, columnA=="xx", columnB %in% c("X","Y"))
mutate(df, columnA_B= A+B)
count(df, columnA, weight=B, sort=T, name='n')
top_n(df,n)
group_by(A,B)
ungroup()
summarise(sumA=sum(A), averageB=mean(B))
# all the above verb can be linked with pipe %>%.
```

## data joining

Join different dataframes or reshape the same one 
```{r}
library(dplyr)
#df_1<A,B,C
#df_2<A,C,C,D

inner_join(x, y, by = c("x_A","y_A"), suffix = c(".x", ".y")) #A,C,C
left_join(x, y,by = c("x_A","y_A"), suffix = c(".x", ".y")) #A,B,C
right_join(x, y, by = NULL, suffix = c(".x", ".y"))#A,C,C,D
full_join(x, y, by = NULL, suffix = c(".x", ".y"))#A,B,C,C,D
semi_join(x, y, by = NULL)#A,C
anti_join(x, y, by = NULL)#B,D

#A semi join differs from an inner join because an inner join will return one row of x for each matching row of y, where a semi join will never duplicate rows of x.

library(tidyr)
pivot_wider(df,id_cols=sampel_id, names_from = OTU, values_from = Abundance)
#turn a long table into a wide table (breakdown OTU, each OTU becomes a new column, the value is the abundance)
pivot_longer(df, cols= 3:7, names_to = "OTU", values_to = Abundance)
pivot_longer(df, cols=starts_with("Day"), names_to="Day_of_treatment",values_to="rank")
```

## apply function to a dataframe

```apply(x, dimension, function)```can apply a function to a dataframe. dimension=1 applies to rows, dimension=2 applies to columns.

```{r}
library(plyr)
apply(df, 2, mean) #same as colMeans(df)
#apply applies to dataframe, needs to specify row or column
lapply(list, sum)
#lapply applies to every element of the list and returns a list
sapply(list, sum)
#sapply also applies to every element of the list, it returns simplified data, e.g., if returns one element each list, it will be turned into a vector instead of a list.
split(df, df$factor_column, drop=F)
#split can split a dataframe into a list of smaller dataframe. Often used with sapply. For example:

t.test_pvalue<- function(df, a,f) {
  l_ttest<-t.test(df[[a]]~df[[f]])
  return(l_ttest[["p.value"]])
}
sapply(split(pd, pd$Source), t.test_pvalue, a="Abundnance",f="Treatment")

ddply(df, .(df$A, df$B), sum)
#split dataframe, apply function and return results in a dataframe.
```

#work with date 

##date is a special data type. str("2020-10-21) is a string, as.Date("2020-10-21") is a date. 
```{r}
library(readr)

Sys.Date() #today's date
as.POSIXct("2020-10-21 11:40:00", tz="America/Chicago")
Sys.time()
Sys.timezone()

library(lubridate)
today()
now()
ymd("2010 Oct 21th")
mdy("10-21-2020")
dmy_hm("21st October 2020 12:00")
parse_date_time("Monday June 1st 2010 at 5pm", order="AmdyIp") #A is week of the day, I indicates 12 hour time, p is pm.
make_date(year="",month="", date="")

#extract partial information
year()
month()
hour()
tz()
day()
yday()
wday()

#a period can also be save and used for calculating dates, eg
period_1<- days(3) + hours(2)
today()+1:10*period_1 #returns a vector of 10, a time series

ymd("2020-01-31") %m+% months(1) # returns 2020-02-29
ymd("2020-01-31") + months(1) #returns NA since there is no 31th in Feb. 

#interval dates
interval(date1, date2) #returns an interval of dates

#Date Stamp 
date_stamp_review <- stamp("The work is reviewed on Saturday, Jan 1, 2000")
date_stamp_review(today())
#will return "The work is reviewed on xxxx (today's date following the above format)

```
