---
title: "Work with dataframe"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## basic characteristics of dataframe
```{r}
library(base)#default loading
rownames()
colnames()
rowSums(x, na.rm=F) 
colSums(x, na.rm=F)
factor(df$column1,ordered=T, levels= c("low", "mid", "high"))
summary(as.factor(df$column1))
```

#piping:select columns, filter rows, add new columns, calculate and summarize data
```{r}
library(dplyr)
arrange(df, A,C,D)
select(df, A:E)
rename(df, A, new_A)
filter(df, columnA=="xx", columnB %in% c("X","Y"))
mutate(df, columnA_B= A+B)
count(df, columnA, weight=B, sort=T, name='n')
top_n(df,n)
group_by(A,B)
ungroup()
summarise(sumA=sum(A), averageB=mean(B))
# all the above verb can be linked with pipe %>%.
```
## data joining

You can include R code in the document as follows:

```{r}
summary(cars)
```

## apply function to a dataframe

You can also embed plots, for example:

```{r }
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
